---
title: "Choropleth with INE data"
author: "Daniel Benedicto Orenes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r lib}
library(leaflet)
library(rgdal)
library(stringr)
library(ggplot2)
library(sp)
library(viridis)
library(htmltools)
```

# Creating an interactive Leaflet map using R and some official INE data (Spain)

To plot geospatial data in R, I choose to use the package leaflet (ggplot and base plot in R also support these files, but they do not render interactive options). One needs two basic things:

1. Data containing the shape of the map you want to plot. For this, there are several options (see packages **sp,sf**). We will use the **sp** package, that deals with *SpatialDataFrames* i.e. objects that are indicated by latitude and longitude coordinates. These can consist on points, lines, or polygones. We will use the polygones options. Each polygone is delimited by a series of lines, and they form a group. From the link below, we will use the spanish map shape file that contains the polygones of the Spanish "Provincias". Shape files for plotting the Spanish map: (http://centrodedescargas.cnig.es/CentroDescargas/catalogo.do?Serie=CAANE)

2. Data to represent in the map, with a proper identifier between the corresponding polygones of the map and the values you want to represent. 

## Leaflet Map

First, let's load and plot a simple leaflet map of Spain. 

```{r simple_map}
shape_file_provincias = "SIANE_CARTO_BASE_S_10M/vigente"
layer_provincias = "se89_10_admin_prov_a_x"
provincias <- readOGR(dsn = shape_file_provincias, layer = layer_provincias)

leaflet(provincias) %>% 
        addPolygons(color = "#444444", weight = 1, opacity = 0.8, fillOpacity = 0.5) 
```

We can check what the **provincias** object looks like

```{r sum}
class(provincias)
summary(provincias)
```


Now, let's add a *fake* column to the data set, in order to assign random colors to each polygon.

```{r color_map}
names = provincias$rotulo
set.seed(123)
rcol = abs(rnorm(length(names),0,4))

# add the data to autonomias data frame
provincias$col = rcol
pal = colorNumeric(topo.colors(10, alpha = 0.7),rcol)

# fill by autonomias
leaflet(provincias) %>% 
        addPolygons(color = "#444444",
                    weight = 1, opacity = 0.8, fill = TRUE,
                    fillColor = ~pal(col),
                    fillOpacity = 0.5)

```

## Data from "padron"

Now, I would like to load some data from the official spanish INE website (add link) that contains data grouped by *provincias*. For instance, we can check some data about population divided by *year*, *provincia* and *gender*. This data comes in a *.px* format. Let's load it. 

```{r padron_data}
library(pxR)
datafile = "cifras_padron/DATOS/padron/l0/2852.px"
data = read.px(datafile, encoding = "latin1")
dp = as.data.frame(data)
# dp is a data frame with 4 columns: Periodo (year), Sexo (gender), 
# Provincias (location), and value (value of the observation). It contains
# 3816 observations, and some 159 NA's which is about 4% of the data. 
summary(dp)
```

Now, let's gather some data to plot. Let's say, we would like to plot,
for the *provincia* of Barcelona, a time series of the population distributed by gender, and also a barplot of the total population by *provincia*, in 2015.

```{r padron_plots}
padronB = dp[grepl("08 Barcelona",dp$Provincias),]
ggplot(padronB[padronB$Sexo %in% c("Hombres","Mujeres"),]) + 
        geom_bar(aes(x = reorder(Periodo,as.numeric(as.character(Periodo)),decreasing = F), y=value, fill = Sexo),
                           stat = "identity",position = position_dodge2()) +
        theme(axis.text.x = element_text("",angle = 90),
              axis.title.x = element_blank()) + 
        ylab("poblacion")


padrontot = dp[dp$Periodo=="2015" & dp$Sexo == "Total" & dp$Provincias != "Total",]
ggplot(padrontot) + geom_bar(stat = "identity",
                             aes(x = reorder(Provincias,value), y = value)) +
        theme(axis.text.x = element_text("",angle = 90), axis.title.x = element_blank()) + 
        ylab("poblacion")
```

## Matching geospatial data with leaflet plot

Now, I would like to plot this data, representing the total population of each *provincia*, on a leaflet map, with a colors scale representing the value of the total population at each provincia.

First we need to match the columns of both data sets. I will adapt the names in INE file to those of the leaflet map. So let's clean up the names in our **padron data frame**, and use the **is.element** function in R to do this. Another approach would be to do a merge using the option *by = identifier*.

```{r clean_names}
pn = padrontot$Provincias
pnames = gsub("^.. ","",pn)
sort(pnames)

# check the match. First, we can observe that they do not have the same length
# I have two more observations in padron than I have in the map. This is due to the 
# canary islands are not included in the map
length(names)
length(pnames)

# we have got 44 matches
length(names[is.element(pnames,names)])
# which are the values that do not match?
pnames[!is.element(pnames,names)]

# From these, Las Palmas y Santa cruz de tenerife we now they are not in the map
# for the others, we need to convert the names in pnames to the corresponding ones in names

# these are the problematic ones
desirednames = names[is.na(charmatch(names,pnames))]
# lets find out the indices in pnames that partially match these
discolos = c("Castell", "Balears", "Alicant","Valencia","Rioja","Coruña")
matchindexdisc = sapply(discolos,function(x) which(grepl(x,pnames)))

# this works, now we got 50 matches
pnames[matchindexdisc]=as.character(desirednames)
length(names[is.element(pnames,names)])


# Let's start subsetting the data frame to those that match before going further
automatch = provincias[names[is.element(names,pnames)],]
automatch$col = padrontot$value[names[is.element(names,pnames)]]
automatch$lab = as.character(padrontot$value[names[is.element(names,pnames)]])
```

Now, we matched the two data frames, and we created a new column in the *map* data frame containing the population. I use a bit of *htmltools* to format the labels of the displaying map. 

```{r leaflet_interactivity}
labels <- sprintf(
        "<strong>%s</strong><br/> Population: %s ",
        automatch$rotulo, automatch$lab) %>% lapply(htmltools::HTML)

# create a color palette matching the values of the population
pal2 = colorNumeric(heat.colors(10),padrontot$value)

# Interactivity with leaflet
leaflet(automatch) %>% 
        addPolygons(color = "#444444",
                    weight = 1, opacity = 0.8, fill = TRUE,
                    fillColor = ~pal2(col),
                    fillOpacity = 0.5,
                    highlight = highlightOptions(
                            weight = 2,
                            color = "red",
                            bringToFront = TRUE),
                    label = labels,
                    labelOptions = labelOptions(style = list("font-weight" = "normal", 
                                                padding = "8px 8px"),
                                                textsize = "15px",direction = "auto")
                    ) %>%
        addLegend(pal = pal2, values = automatch$col, opacity = 0.7, title = "Population",
                  position = "bottomleft")


```








